---
title: "disaster"
author: "Mingrui Du"
date: "2023-10-30"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#install.packages("rfema", repos = "https://ropensci.r-universe.dev")
library(httr)
library(jsonlite)
library(rjson)
library(rfema)
suppressMessages(library(dplyr))
library(tidyverse)
library(stringr)
library(maps)
suppressMessages(library(plotly))
suppressMessages(library(usmap))
# library(ggmap)
```

## Data acquisition and assessment

### Flood Data

Two data sets obtained from [Federal Emergency Management Agency (FEMA)](https://www.fema.gov/about/openfema/data-sets), v1 containing financial assistance and v2 for all declared disasters. 

[v1 (Fema Web Disaster Summaries)](https://www.fema.gov/openfema-data-page/fema-web-disaster-summaries-v1) is a 14-column, formatted data frame containing: sequential disaster number; total approved number of Individual Assistance (IA); approved amount($) for the Individual and Households Program (IHP), which includes Housing Assistance (HA) and Other Needs Assistance (ONA); amount of Public Assistance grant funding (PA) and [Hazard Mitigation Grant Program (HMGP)](https://www.fema.gov/grants/mitigation/hazard-mitigation). 

[v2 (Disaster Declarations Summaries)](https://www.fema.gov/openfema-data-page/disaster-declarations-summaries-v2) contains disaster time, counties, incident type, disaster descriptions and assistance programs. Among these, column "declarationTitle" is a concatenation of all incidents included in one declared disaster. 

Since the assignment focuses on year 2020-2021, the two data set will be filtered respectively for further investigation. 

### CENSUS data

The three files of CENSUS data posted on bb are originally obtained from [U.S. Census Bureau](https://www.census.gov/data/developers/data-sets/acs-5year.html), while [S1701](https://www.census.gov/acs/www/data/data-tables-and-tools/subject-tables/) stands for poverty status in the past 12 Months, [DP05](https://data.census.gov/table/ACSDP5Y2021.DP05) for demographic and housing estimates, [B25001](https://data.census.gov/table?q=B25001) for housing units. 

### NOAA



## Questions

当地受灾情况与房屋数量关系: FEMA、CENSUS


## Coding

### Data Loading

Data sets are read via local files and api. 

```{r}
## via API
## Tutorial: https://www.fema.gov/about/openfema/api
base_fema <- 'https://www.fema.gov/api/open/'
version_2 <- 'v2/'
entity <- 'DisasterDeclarationsSummaries'
## select: 2020-2021, flood
info_key <- "?$filter=incidentBeginDate%20ge%20'2020-01-01'%20and%20incidentBeginDate%20lt%20'2022-01-01'%20and%20incidentType%20eq%20'Flood'"
url_fema <- paste0(base_fema, version_2, entity, info_key)
rawd <- GET(url_fema)  
raw_text <- content(rawd, as = "text", encoding = "UTF-8")
df <- fromJSON(raw_text, flatten = TRUE)
fema_v2 <- data.frame(df$DisasterDeclarationsSummaries)

## via local files
fema_v1 <- read.csv("FemaWebDisasterSummaries.csv", header = T)

## S1701: Poverty
pov_20 <- read.csv("ACSST5Y2020.S1701-Data.csv", header = T)
pov_20 <- pov_20 %>% select_if(~ length(unique(.)) > 2)
## all.equal(colnames(pov_20), colnames(pov_21))

pov_21 <- read.csv("ACSST5Y2021.S1701-Data.csv", header = T)
pov_21 <- pov_21 %>% select_if(~ length(unique(.)) > 2)


## DP05: ACS DEMOGRAPHIC AND HOUSING ESTIMATES
demo_20 <- read.csv("ACSDP5Y2020.DP05-Data.csv", header = T)
demo_20 <- demo_20 %>% select_if(~ length(unique(.)) > 2)
# demo_20 <- demo_20[, grepl("E$",colnames(demo_20))]
# colnames(demo_20) <- unlist(demo_20[1, ])
# demo_20 <- demo_20[-1,]
#### preparation for splitting columns

# colnames(demo_20) <- sapply(colnames(demo_20), function(x) gsub("Estimate!!SEX AND AGE!!Total population", "SexAge_", x))
# colnames(demo_20) <- sapply(colnames(demo_20), function(x) gsub("Estimate!!RACE!!Total population", "RaceTotal_", x))
# colnames(demo_20) <- sapply(colnames(demo_20), function(x) gsub("RaceTotal_!!One race!!American Indian and Alaska Native", "IndianAlaska_", x))
# colnames(demo_20) <- sapply(colnames(demo_20), function(x) gsub("RaceTotal_!!One race!!Asian!!", "Asian_", x))
# colnames(demo_20) <- sapply(colnames(demo_20), function(x) gsub("RaceTotal_!!One race!!Native Hawaiian and Other Pacific Islander!!", "HwiPacific_", x))



##colnames(demo_20) <- sapply(colnames(demo_20), function(x) gsub("_!!", "_", x))


demo_21 <- read.csv("ACSDP5Y2021.DP05-Data.csv", header = T)
demo_21 <- demo_21 %>% select_if(~ length(unique(.)) > 2)
# colnames(demo_21) <- unlist(demo_21[1, ])
# demo_21 <- demo_21[-1,]


## B25001: Housing Units
haus_uni_20 <- read.csv("ACSDT5Y2020.B25001-Data.csv", header = T)
haus_uni_20 <- haus_uni_20 %>% select_if(~ length(unique(.)) > 2)
colnames(haus_uni_20) <- unlist(haus_uni_20[1, ])
haus_uni_20 <- haus_uni_20[-1,]

haus_uni_21 <- read.csv("ACSDT5Y2021.B25001-Data.csv", header = T)
haus_uni_21 <- haus_uni_21 %>% select_if(~ length(unique(.)) > 2)
colnames(haus_uni_21) <- unlist(haus_uni_21[1, ])
haus_uni_21 <- haus_uni_21[-1,]




```
### Data Cleaning

FEMA v1

```{r}
## v1
fema_v1 <- fema_v1 %>%
  select(-hash, -lastRefresh, -id)
## select data by disaster number in v2: 
fema_v1 <- fema_v1 |> 
  semi_join(fema_v2, by = "disasterNumber")
```

FEMA v2: remove unused columns, split BeginDate, Area, DeclarationTitle then separate by year

```{r}
## v2
fema_v2 <- fema_v2 %>%
  select(-hash, -lastRefresh, -id, -femaDeclarationString) %>% 
  select(iaProgramDeclared, where(~length(unique(.)) > 1))
## 'iaProgramDeclared' in v2 is a single value col 
## but I want to save it as a funding indicator

fema_v2$County <- 

## Split concatenation columns:
### incidentBeginDate into Year, Month, Date
fema_v2$incidentBeginDate <- as.character(fema_v2$incidentBeginDate)
fema_v2$incidentBeginDate <- gsub("T00:00:00.000Z", "", fema_v2$incidentBeginDate)
fema_v2 <- fema_v2 |>
                separate_wider_delim(cols = incidentBeginDate,
                                     delim = "-",
                                     names = c("Year", 
                                               "Month",     
                                               "Date"))

### declarationTitle
fema_v2$declarationTitle <- gsub(", AND", ",", fema_v2$declarationTitle)
fema_v2$declarationTitle <- gsub("AND ", ", ", fema_v2$declarationTitle)
fema_v2$declarationTitle <- gsub("SEVERE, STORMS", "SEVERE STORMS", fema_v2$declarationTitle)
### unique(grepl(" AND ", fema_v2$declarationTitle))
fema_v2 <- fema_v2 |>
                separate_wider_delim(cols = declarationTitle,
                                     delim = ",",
                                     names = c("temp1", 
                                               "temp2",     
                                               "temp3",
                                               "temp4",
                                               "temp5"),
                                     too_many = "error",
                                     too_few = "align_start")
fema_v2 <- fema_v2 %>% mutate_at(vars(temp1:temp4), ~trimws(.))

## create FIPS code column
fema_v2$fips <- paste0(fema_v2$fipsStateCode, fema_v2$fipsCountyCode, sep = "")
fema_v2$fips <- ifelse(grepl("000", fema_v2$fips), NA, fema_v2$fips)


## Separate FEMA v2 by year 2020 and 2021
fema_v2_20 <- fema_v2 %>% filter(Year == '2020')
fema_v2_21 <- fema_v2 %>% filter(Year == '2021')

```

CENSUS: Split Census data sets by Estimates and Percent

```{r}
## census data
## Poverty
Area <- as.data.frame(pov_20$NAME)|>
    separate_wider_delim(cols = `pov_20$NAME`,
                         delim = ",",
                         names = c("County","State"),
                         too_few = "align_start",
                         too_many = "error")
Area[1,1] <- "County"
Area[1,2] <- "State"
Area$County <- trimws(Area$County)
Area$State <- trimws(Area$State)
## Divide Poverty Data
pov_20 <- pov_20[, grepl("E$",colnames(pov_20))] ## Remove all non-estimate columns 
pov_20_totl <- data.frame(Area, pov_20[ ,grepl("C01", colnames(pov_20))]) ## Total 
pov_20_blo <- data.frame(Area, pov_20[ ,grepl("C02", colnames(pov_20))]) ## Below Poverty Level
pov_20_blopc <- data.frame(Area, pov_20[ ,grepl("C03", colnames(pov_20))]) ## Percent of Blo

## Replace col names by first row
colnames(pov_20_totl) <- unlist(pov_20_totl[1, ])
pov_20_totl <- pov_20_totl[-1,]

## age
colnames(pov_20_totl) <- sapply(colnames(pov_20_totl), function(x) gsub("Population for whom poverty status is determined!!AGE!!", "age_", x))
colnames(pov_20_totl) <- sapply(colnames(pov_20_totl), function(x) gsub("Under 18 years!!", "", x))
colnames(pov_20_totl) <- sapply(colnames(pov_20_totl), function(x) gsub("18 to 64 years!!", "", x))

## Education
colnames(pov_20_totl) <- sapply(colnames(pov_20_totl), function(x) gsub("Population for whom poverty status is determined!!EDUCATIONAL ATTAINMENT!!", "edu_", x))
colnames(pov_20_totl) <- sapply(colnames(pov_20_totl), function(x) gsub("Population 25 years and over!!", "", x))

## Employment
colnames(pov_20_totl) <- sapply(colnames(pov_20_totl), function(x) gsub("Population for whom poverty status is determined!!EMPLOYMENT STATUS!!", "emp_", x))
colnames(pov_20_totl) <- sapply(colnames(pov_20_totl), function(x) gsub("Civilian labor force 16 years and over!!", "", x))


```


split concatenation columns

```{r}


```

```{r}
## CENSUS
cty_state <- function(c){
  c <- c |> separate_wider_delim(cols = `Geographic Area Name`,
                                 delim = ",",
                                 names = c("County", 
                                           "State"))
  c$County <- trimws(c$County)
  c$State <- trimws(c$State)
  return(c)
}
pov_20 <- cty_state(pov_20); pov_21 <- cty_state(pov_21)
#demo_20 <- cty_state(demo_20); demo_21 <- cty_state(demo_21)
haus_uni_20 <- cty_state(haus_uni_20); haus_uni_21 <- cty_state(haus_uni_21)
haus_uni_20$County <- gsub(" County", "", haus_uni_20$County)
haus_uni_21$County <- gsub(" County", "", haus_uni_21$County)


```

## EDA

```{r}
## 1 fema & housing unit

maps::county.fips %>%
  as.tibble %>% 
  extract(polyname, c("region", "subregion"), "^([^,]+),([^,]+)$") ->
  dfips

map_data("county") %>% 
  left_join(dfips) ->
  dall

dall %>% 
  mutate(is_example = fips %in% fema_v2$fips) %>% 
  ggplot(aes(long, lat, group = group)) +
  geom_polygon(aes(fill=is_example), color="gray70") +
  coord_map()
```


## References

[R FEMA](https://docs.ropensci.org/rfema/#installation)

[NOAA Storm Events Database](https://www.ncdc.noaa.gov/stormevents/ftp.jsp)

[OpenFEMA Data Sets](https://www.fema.gov/about/openfema/data-sets)

[API in R](https://statisticsglobe.com/api-in-r)

[ggmap](https://mapsplatform.google.com/)
